---
title: "Homework 6:Data wrangling wrap up"
theme: yeti
output:
  github_document:
    toc: yes
    df_print: paged
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Character Data
Loading libraries:
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(stringr))
```

#### Basics
```{r}
simpleStr <- "Simple string"
strWithQuotes <- 'Note I\'m using single quotes and escape chars to \"make it fancy\"\nNew line added above'
writeLines(strWithQuotes) # to write clean version - without escapes
```

Simple operations
```{r}
?str_c
str_length(simpleStr)
str_c("These strings", "are actually combined", "by a comma", sep=", ")
str_c(c("These strings", "are actually combined", "by a comma"), collapse = ", ") # easier to see
# str_c is verctorised, so we can add prefix, and suffix for a vector as below
str_c("beforeStr ", c("a", "b", "c"), " afterStr")
```

Substringing is very useful
```{r}
weekDays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
str_sub(weekDays, 1, 3) 
# negative numbers count backwards from end
str_sub(weekDays, -3, -1)
# we can also modify first element of every item in the vector
str_sub(weekDays, 1, 1) <- str_to_lower(str_sub(weekDays, 1, 1))
weekDays
```

Using locale for sorting. 

For more information refer to [Iso list](https://en.wikipedia.org/wiki/List_of_ISO_639-1_codes)
```{r}
str_sort(weekDays, locale = "haw") # Hawaiian
str_sort(weekDays, locale = "en") # English
```

Working through exercises:

In code that doesnt use stringr, youll often see paste() and paste0(). Whats the difference between the two functions? What stringr function are they equivalent to? How do the functions differ in their handling of NA?

Answer: The difference between `paste()` and `paste0()` is that the argument sep by default is " " (paste) and "" (paste0). `paste0()` is faster than `paste()` if our objective is to concatenate strings without spaces because we don't have to specify the argument sep
```{r}
?paste
paste(c("aaa", "bb"), sep = "", collapse = "," )
?paste0
paste(c("aaa", "bb"),  collapse = ", " ) # same output
```

In your own words, describe the difference between the sep and collapse arguments to str_c().

`sep` is a character string to separate the terms while collapse is an optional character string to separate the results.

Use str_length() and str_sub() to extract the middle character from a string. What will you do if the string has an even number of characters?
```{r}
s <- "abc"
str_sub(s, str_length(s)/2+1, str_length(s)/2+1) # since 0 based
s_even <- "abcd" # middle is b, breaking ties here between b and c
str_sub(s_even, str_length(s)/2+1, str_length(s)/2+1) # since 0 based
```

What does str_wrap() do? When might you want to use it?
Answer: You use it when you need to format some text in a nice manner, for ease of perusing through it.
```{r}
?str_wrap
```

What does str_trim() do? Whats the opposite of str_trim()?
Answer: We use it when we want to remove whitespace.  `str_pad()` is it's opposite
```{r}
?str_pad
```

Small demo:
```{r}
?str_trim
dirtySentence <- "     No spaces in front/end of string.       "
str_trim(dirtySentence)
```

#### Regular expressions
```{r}
# Usage of '.': Meaning any one character
str_view(c("Monday", "Tuesday"), ".day")
str_view(c("Mon.day", "Tuesday"), "\\.day") # escape character for . itself!
str_view(c("lay\\lom"), ".\\\\") # Escape for '\' character itself!
```

Matching more complicated sequences like `"'\`:
```{r}
complicatedStr <- 'something\"\'\\something'
writeLines(complicatedStr)
str_view(complicatedStr, "\"\\'\\\\")
```

Exercises:
Explain why each of these strings don't match a `\`: `\`, `\\`, `\\\`.
Answer: In short, you need an escape `\` for the escape `\\` of this character `\`. So at least 4 `\` chars.

How would you match the sequence `"'\`?
```{r}
m <- '\"\'\\\ something'
writeLines(m)
str_view(m, '\\"\'\\\\')
```

What patterns will the regular expression `\..\..\..` match? How would you represent it as a string?
```{r}
# This is anything like so: \fg\hj\ki
pattern <- "\\..\\..\\.." # pattern for these strings
```

Anchors for making the regular expression so that it matches from the start or end of the string.
```{r echo=FALSE}
anchorStr <- c("all", "starting", "with", "a", "ending", "with", "h")
str_view(anchorStr, "^a")
str_view(anchorStr, "h$")
```

Exercises:
Matching literal string `$^$`
```{r}
litStr <- "before$^$after"
str_view(litStr, "\\$\\^\\$")
```

Given the corpus of common words in `stringr::words`, create regular expressions that find all words that:

Start with "a".
```{r}
# \\w is short hand for [A-Za-z0-9_]
pat <- 'a\\w*\\b'
str_view(words %>% head(), pat)
```

End with ???t???
```{r}
pat <- '[A-Za-z0-9_]*t$'
str_view(words %>% head(), pat)
```

Are exactly four letters long. (Dont cheat by using str_length()!)
Have seven letters or more.
```{r}
pat_4 <- '^....$'
str_view(words %>% head(), pat_4)
pat_7plus <- "^.......+$"
str_view(words %>% head(), pat_7plus)
```


There are a number of special patterns that match more than one character. You've already seen `.`, which matches any character apart from a newline. There are four other useful tools:

* `\d`: matches any digit.
* `\s`: matches any whitespace (e.g. space, tab, newline).
* `[abc]`: matches a, b, or c.
* `[^abc]`: matches anything except a, b, or c.

Therefore we need to be careful and excape the `\d`. 
Star `*` symbol is special for regex, it means 0 or more instances of previous character, but using `[]` means literally match all characters within the brackets.
```{r}
abcRegex <- c("abc", "a.c", "a*c", "a c")
str_view(abcRegex, "a[*]c")
str_view(abcRegex, "a[.]c")
str_view(abcRegex, ".[ ]c")
str_view(c("color", "colour"), "colo(|u)r") # picks both due to usage of `|` operator
```

How many times a pattern matches is defined by:

* `?`: 0 or 1
* `+`: 1 or more
* `*`: 0 or more

To specify the number of matches precisely:

* `{n}`: exactly n
* `{n,}`: n or more
* `{,m}`: at most m
* `{n,m}`: between n and m
```{r}
matchings <- "MDCCCLXXXVIII is 1888 in roman numerals"
str_view(matchings, "CC?")
str_view(matchings, "CC+")
str_view(matchings, 'C[LX]+') # will match CLXXX
str_view(matchings, "C{2}") # will match only CC
str_view(matchings, "C{2,}") # will match 2 or more Cs
```

We can also backreference as so:
```{r}
# Find any repetition of any two letters
str_view(c("ananas", "banana", "apapa", "aaaaa"), "(..)\\1", match = TRUE)

```

#### Tools
To determine if a character vector matches a pattern, we use str_detect(). 
It returns a logical vector the same length as the input:
```{r}
repStr <- c("ananas", "banana", "apapa", "aaaaa")
str_detect(repStr, "w") # none has 'w'
str_detect(repStr, "a") # they all have 'a'
```

```{r}
# How many weekday names start with t?
sum(str_detect(weekDays, "^t")) 
```

Playing with vowels and consonants:
```{r}
# Find all words containing at least one vowel, and negate
no_vowels_1 <- !str_detect(c("abbb", "eddd"), "[aeiou]")
# Find all words consisting only of consonants (non-vowels)
no_vowels_2 <- str_detect(c("abbb", "addd"), "^[^aeiou]+$")
identical(no_vowels_1, no_vowels_2)
```

Matching patterns can be done in 2 ways - logical subsetting(1) or `str_subset` wrapper(2) :
```{r}
words[str_detect(words, "x$")] #(1)
str_subset(words, "x$") #(2)
```

A variation on `str_detect()` is `str_count()`: rather than a simple yes or no, it tells you how many matches there are in a string:

```{r}
fruitNames <- c("apple", "banana", "pear")
str_count(fruitNames, "a")

# On average, how many vowels per word?
mean(str_count(words, "[aeiou]"))
```

We can also use `mutate` as so, to get matches in a tibble for further usage:
```{r}
df <- tibble(
  word = fruitNames,
  i = seq_along(word)
)

df %>%
  mutate(
    vowels = str_count(word, "[aeiou]"),
    consonants = str_count(word, "[^aeiou]")
  )
```

Matches are not overlapping: 
```{r}
str_count("abababa", "aba")
str_view_all("abababa", "aba")
```

If we have a corpus, and want to identify all sentences containing a color we could do this:
```{r}
# Notice this will not include words like "flickered"!
colours <- c(" red", " orange", " yellow", " green", " blue", " purple")
allColours <- str_c(colours, collapse = "|")
manyColorSent <- sentences[str_count(sentences, allColours) > 1]
str_view_all(manyColorSent, allColours)
str_extract(manyColorSent, allColours) # which colors matched from above?
str_extract_all(manyColorSent, allColours, simplify = TRUE) # exactly which ones, for each sentence?
```

From the Harvard sentences data, let's extract:

The first word from each sentence.
```{r}
head(sentences)
firstWords <- str_extract(sentences, "([A-Za-z]+)")
head(firstWords)
```

All words ending in ing.
```{r}
allWords <- str_extract_all(sentences, boundary("word")) # take all words
allWordsList <- unlist(allWords) # make a char vector
ingWords <- str_match_all(allWordsList, ".+ing") # only ing ending words
unlist(ingWords)
```

Matching `a ___` and `the ___`:
```{r}
noun <- "(a|the) ([^ ]+)"
has_noun <- sentences %>%
  str_subset(noun) %>%
  head(10)
# To get a vector
has_noun %>% 
  str_extract(noun)
# To get a matrix
has_noun %>% 
  str_match(noun)
```

Replacing vowels:
```{r}
fruitNames <- c("apple", "pear", "banana")
str_replace(fruitNames, "[aeiou]", "-") # replaces first vowel
str_replace_all(fruitNames, "[aeiou]", "-") # replaces all vowels
```

Using backreferences:
```{r}
head(sentences)
# First 3 words of each sentence have twisted order now.
sentences %>% 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>% 
  head(6)
```

Splitting:
```{r}
sentences %>%
  head(5) %>% 
  str_split(" ", simplify = TRUE)
str_view_all(sentences %>% head(1), boundary("word"))
```

#### Other types of patterns

We can ignore case for matching patterns
```{r}
apples <- c("apple", "APPLE")
str_view(apples, regex("apple", ignore_case = TRUE))
```

Recognizing phone numbers:
```{r}
phone <- regex("
  \\+?                   # beginning +
  (\\d{1,4})? # country code
  [-]?                # optional closing parens, space, or dash
  \\(?                  # optional opening parens
  (\\d{3})              # area code
  [) -]?                # optional closing parens, space, or dash
  (\\d{3})              # another three numbers
  [ -]?                 # optional space or dash
  (\\d{4})              # three more numbers
  ", comments = TRUE)

str_match("604-908-8172", phone)
str_match("+1-604-908-8172", phone)
str_match("+12344-604-908-8172", phone) # erroneous, accepts only 2344
```

#### More reg ex

`apropos()` searches all objects available from the global environment. 
This is useful to better remember the name of the function.
```{r}
apropos("replace")
```

`dir()` lists all the files in a directory. 
The `pattern` argument takes a regular expression and 
only returns file names that match the pattern.
For example, you can find all the R Markdown files in the current directory with:
```{r}
head(dir(pattern = "\\.Rmd$"))
```

#### stringi
Using `stringi` to generate random text
```{r}
# make 10, length 2, text!
stringi::stri_rand_strings(10, 2, pattern = "[A-Za-z0-9]")
```

# 2. Work with a list
This tutorial is inspired by [this blogpost](http://varianceexplained.org/r/trump-tweets/). 

## Trump Android words

```{r}
library(purrr)
suppressMessages(library(dplyr))
library(tibble)
```

Lets load tweets from Trump's account.
```{r}
load(url("http://varianceexplained.org/files/trump_tweets_df.rda"))
tweets <- trump_tweets_df$text
tweets %>% head(3) %>% strtrim(70)
```

Our mission: extract the following Trump Android words from the tweets.
```{r}
# Define Trump Android words!
regex <- "badly|crazy|weak|spent|strong|dumb|joke|guns|funny|dead"
# Chose only certain tweets that have necessary complexity
tweets <- tweets[c(1, 2, 5, 6, 198, 347, 919)]
tweets %>% strtrim(70)
```

#### Using `gregexpr()`
The base function `gregexpr()` is used to locate all the Trump Android words inside the tweets. 

```{r}
(matches <- gregexpr(regex, tweets))
# matches is a list, one element per each element of tweets
# Each element is an integer vector.
#   Its -1 if no matches found.
#   Holds the position(s) of the first character of each match, otherwise.
# Each element has two attributes. Consider match.length. Let us not speak of the other one.
#   Its -1 if no matches found.
#   Holds the length(s) of each match, otherwise.
m <- matches[[7]]
```

Let's get to know this awkward list:
```{r}
attr(m, which = "match.length")
```

#### `purrr`ing data!

Putting that into purrr:
```{r}
lens <- function(x) attr(x, which = "match.length")
map(matches, lens)
# or 
(match_length <- map(matches, attr, which = "match.length"))
```

The above list con
Let's compute how many Trump Android words appear in each tweet.
```{r}
m <- matches[[1]]
sum(m > 0)
m <- matches[[7]]
sum(m > 0)
```


```{r}
androidWords <- function(x) sum(x > 0)
map(matches, androidWords) # the result is a list
map_int(matches, androidWords) # the result is a vector
```

Confirm that this is, indeed, different from just taking the lengths of the elements of `matches`:
```{r}
tibble(
  naive_length = lengths(matches),
  n_words = map_int(matches, androidWords)
)
```

Let's create an almost-copy of matches and call it `match_first`. How will it differ? Remove the attributes from the elements of matches, so there's less clutter when we print.

Hint: `as.vector()` will strip attributes.
```{r}
(match_first <- map(matches, as.vector))
```

Let's check the 7th sentence. 
```{r}
(tweet <- tweets[7]) # get the tweet
(t_first <- match_first[[7]]) # get the indexes of the first match of each android word appearing
(t_length <- match_length[[7]]) # get lengths 
(t_last <- t_first + t_length - 1) # compute end of android word
substring(tweet, t_first, t_last) # substring them, to see which ones they are!
```

And now the first tweet. This shouldn't have any android word.
```{r}
(tweet <- tweets[1])
(t_first <- match_first[[1]])
(t_length <- match_length[[1]])
(t_last <- t_first + t_length - 1)
substring(tweet, t_first, t_last)
```

This is how we did the above exercises. Let's put these into purrr. 
```{r}
# first  <- an element of matches
# length <- an element of match_length
# last   <- first + length - 1
(match_last <- map2(match_first, match_length, ~ .x + .y - 1))
```

This is all put together, the real magic of `purrr`!
```{r}
pmap(list(text = tweets, first = match_first, last = match_last), substring)
```

#### Concise summary of work done.

Summary of everything we did in this part of the homework, being as concise as possible!
```{r}
tibble(text = tweets,
       first = gregexpr(regex, tweets)) %>% 
  mutate(match_length = map(first, ~ attr(.x, which = "match.length")),
         last = map2(first, match_length, ~ .x + .y - 1)) %>%
  select(-match_length) %>% 
  pmap(substring)
```

Thanks for looking at my homework. Cheers!
